---
title: "Choosing Lambda in Mixed-Subjects IRT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing Lambda in Mixed-Subjects IRT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

The mixed-subjects estimator fits a 2PL calibration by minimizing

```{r objective, echo = FALSE, results = "asis"}
cat("`L_human + lambda * (L_generated - L_paired_llm)`")
```

The tuning parameter `lambda` controls how strongly generated LLM responses are
used after subtracting the paired LLM correction. Setting `lambda = 0` gives the
human-only expected-count calibration. Larger values use more information from
the generated response sample, but they can also amplify LLM bias when the
paired correction is weak.

This vignette distinguishes three package workflows:

- `diagnose_lambda_grid()`: a sensitivity diagnostic. It is not a valid final
  tuning rule.
- `tune_lambda_ability()`: chooses a lambda by minimizing propagated
  ability-score risk on a target scoring population.
- `tune_lambda_ability_crossfit()`: estimates lambda out of sample by fold, then
  fits the final split-sample estimator with fold-specific lambdas.

## Example data

The example uses simulated human responses, paired LLM responses for the same
human rows, and additional generated LLM responses. In an applied project,
replace these matrices with your study data.

```{r data}
library(mixedsubjectsirt)

set.seed(2027)

n_human <- 120
n_generated <- 350
n_items <- 5

human_pars <- data.frame(
  item = paste0("Item", seq_len(n_items)),
  a = c(0.8, 1.0, 1.2, 1.35, 1.55),
  d = c(-0.8, -0.3, 0.1, 0.45, 0.9)
)
human_pars$b <- -human_pars$d / human_pars$a

llm_pars <- human_pars
llm_pars$a <- pmax(0.35, 0.9 * human_pars$a)
llm_pars$d <- human_pars$d + 0.25
llm_pars$b <- -llm_pars$d / llm_pars$a

theta_human <- rnorm(n_human)
observed <- simulate_2pl(theta_human, human_pars)
predicted <- simulate_2pl(theta_human, llm_pars)
generated <- simulate_2pl(rnorm(n_generated), llm_pars)

lambda_grid <- c(0, 0.25, 0.5, 0.75, 1)
```

The examples below use `human_pars` as `initial_pars` so the vignette runs
quickly. In a real package workflow, a typical starting point is
`initial_pars = fit_2pl(observed)$pars`.

## Fixed Lambda

When `lambda` is fixed by design, call `fit_mixed_subjects()` directly.

```{r fixed-lambda}
fixed_fit <- fit_mixed_subjects(
  observed = observed,
  predicted = predicted,
  generated = generated,
  lambda = 0.5,
  initial_pars = human_pars,
  n_quad = 7,
  control = list(maxit = 100)
)

fixed_fit$item_pars
```

Use this route when a simulation study, design analysis, or preregistered plan
already specifies `lambda`.

## Sensitivity Diagnostics

`diagnose_lambda_grid()` fits candidate lambda values and reports the mixed
objective and the observed-human expected-count loss. This is useful for seeing
how sensitive the fitted calibration is to lambda.

```{r diagnostic}
diagnostic <- diagnose_lambda_grid(
  lambda_grid = lambda_grid,
  observed = observed,
  predicted = predicted,
  generated = generated,
  initial_pars = human_pars,
  n_quad = 7,
  control = list(maxit = 100)
)

diagnostic$summary
diagnostic$lowest_observed_loss_lambda
```

The value in `lowest_observed_loss_lambda` is not a final PPI++ tuning rule. It
is the lambda with the smallest observed-human expected-count loss after fitting.
That criterion can overfit the labeled sample and is not the same as minimizing
downstream ability-score uncertainty.

## Ability-Risk Tuning

For scoring applications, `tune_lambda_ability()` targets the quantity

```{r risk-formula, echo = FALSE, results = "asis"}
cat("`mean(g_i' Sigma_lambda g_i)`")
```

where `Sigma_lambda` is the full sandwich covariance matrix of the item
parameters and `g_i` is the gradient of the estimated ability score for response
pattern `i` with respect to those item parameters.

```{r ability-tuning}
ability_tuned <- tune_lambda_ability(
  lambda_grid = lambda_grid,
  observed = observed,
  predicted = predicted,
  generated = generated,
  target_resp = observed,
  initial_pars = human_pars,
  n_quad = 7,
  control = list(maxit = 100)
)

ability_tuned$summary
ability_tuned$best_lambda
```

The selected calibration is available as `ability_tuned$best_fit`.

```{r ability-fit}
ability_tuned$best_fit$item_pars
```

You can also inspect the covariance and ability-risk components directly.

```{r covariance-risk}
Sigma <- vcov(ability_tuned$best_fit)
dim(Sigma)

risk <- ability_risk(
  resp = observed,
  fit_or_pars = ability_tuned$best_fit,
  vcov = Sigma
)

risk$summary
```

In simulation studies, pass `theta_true` to include squared ability-estimation
error in addition to propagated item-parameter uncertainty.

```{r ability-tuning-truth}
ability_tuned_with_truth <- tune_lambda_ability(
  lambda_grid = lambda_grid,
  observed = observed,
  predicted = predicted,
  generated = generated,
  target_resp = observed,
  theta_true = theta_human,
  initial_pars = human_pars,
  n_quad = 7,
  control = list(maxit = 100)
)

ability_tuned_with_truth$summary
```

For real data, `theta_true` is normally unavailable. The default risk is then
the propagated calibration component, `g_i' Sigma_lambda g_i`.

## Cross-Fitted Ability-Risk Tuning

`tune_lambda_ability()` uses the same labeled responses to tune and fit. That is
useful for prototyping, but final finite-sample analyses should avoid using a
fold's labels to tune the lambda applied to that fold's paired correction.

`tune_lambda_ability_crossfit()` estimates lambda on training folds and then
fits the final split-sample estimator with the resulting fold-specific lambdas.

```{r crossfit}
split_id <- rep(1:2, length.out = nrow(observed))

crossfit_tuned <- tune_lambda_ability_crossfit(
  lambda_grid = c(0, 0.5, 1),
  observed = observed,
  predicted = predicted,
  generated = generated,
  split_id = split_id,
  initial_pars = human_pars,
  n_quad = 7,
  control = list(maxit = 100)
)

crossfit_tuned$lambda_by_split
crossfit_tuned$final_fit
```

The final fit stores both the fold-specific lambdas and their split-size
weighted mean for the generated term.

```{r crossfit-fit}
crossfit_tuned$final_fit$split$lambda_by_split
crossfit_tuned$final_fit$lambda_generated
```

## Choosing a Procedure

Use a fixed lambda when the value is determined by design or by a simulation
study. Use `diagnose_lambda_grid()` to understand sensitivity, not to make a
final inferential choice. Use `tune_lambda_ability()` when the practical target
is ability scoring and you want the lambda that minimizes propagated scoring
risk on a chosen target population. Use `tune_lambda_ability_crossfit()` for the
same target when finite-sample adaptivity matters and you want lambda estimated
out of sample.

The target population matters. `target_resp = observed` tunes for the observed
human response patterns. In operational scoring, you may prefer a larger target
matrix representing the population that will actually be scored.
