---
title: "Choosing Lambda in Mixed-Subjects IRT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing Lambda in Mixed-Subjects IRT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

The mixed-subjects estimator calibrates a 2PL model by minimizing a
PPI++-style combined loss over observed human responses, paired LLM responses,
and additional generated LLM responses.

This vignette explains how to choose $\lambda$ and which functions to use. The
short answer:

| Task | Function |
|------|----------|
| Theoretical diagnostic | `tune_lambda_ppi_score()` |
| **Recommended practical tuning** | **`tune_lambda_ability_risk(..., fit_fn = fit_mixed_subjects_mml)`** |
| Frozen EC (fast approximation) | `tune_lambda_ability_risk()` (default `fit_fn`) |
| Cross-fitted tuning | `tune_lambda_ability_risk_crossfit()` |
| Per-item tuning (experimental) | `tune_lambda_ability_risk_item()` |

## Two objectives, two estimators

**Why there are two lambda objectives:**

- `tune_lambda_ppi_score()` minimizes $\text{Tr}(\Sigma_\gamma)$ — the trace of
  the item-parameter covariance matrix (PPI++ Proposition 2). This is a
  *theoretical diagnostic*: it measures whether the LLM prediction reduces
  item-parameter variance. It answers "is the predictor statistically useful?"

- `tune_lambda_ability_risk()` minimizes $E[g'\Sigma_\gamma g]$ — the propagated
  ability-score risk. This is the *practical criterion*: it answers "does using
  the LLM improve downstream test scoring?" Use this for operational calibration.

**Why there are two estimators:**

The original `fit_mixed_subjects()` uses frozen expected-count posteriors. This
creates a gradient asymmetry when the LLM item parameters differ from human
parameters, systematically inflating discriminations and driving
`tune_lambda_ability_risk` to select $\lambda = 0$ even when the LLM is
genuinely informative.

`fit_mixed_subjects_mml()` recomputes posteriors at every gradient evaluation,
eliminating this asymmetry. Its sandwich covariance (`vcov_mixed_subjects_mml`,
called automatically via `vcov()`) uses Louis' (1982) observed-information
correction for the bread, giving honest uncertainty estimates. Use the MML
estimator unless speed is a binding constraint.

## Example data

```{r data}
library(mixedsubjectsirt)

set.seed(2027)

n_human     <- 120
n_generated <- 350
n_items     <- 5

human_pars <- data.frame(
  item = paste0("Item", seq_len(n_items)),
  a    = c(0.8, 1.0, 1.2, 1.35, 1.55),
  d    = c(-0.8, -0.3, 0.1, 0.45, 0.9)
)
human_pars$b <- -human_pars$d / human_pars$a

llm_pars   <- human_pars
llm_pars$a <- pmax(0.35, 0.9 * human_pars$a)
llm_pars$d <- human_pars$d + 0.25
llm_pars$b <- -llm_pars$d / llm_pars$a

theta_human <- rnorm(n_human)
observed    <- simulate_2pl(theta_human, human_pars)
predicted   <- simulate_2pl(theta_human, llm_pars)
generated   <- simulate_2pl(rnorm(n_generated), llm_pars)

lambda_grid <- c(0, 0.25, 0.5, 0.75, 1)
```

The examples use `human_pars` as `initial_pars` for speed.

## Step 1: PPI++ score diagnostic

`tune_lambda_ppi_score()` estimates the PPI++ Proposition 2 lambda using the
same human posterior weights for both human and paired-LLM score vectors.
F = Y (identical predictions) gives exactly $N/(n+N)$.

```{r ppi-score}
ppi_score <- tune_lambda_ppi_score(
  observed    = observed,
  predicted   = predicted,
  item_pars   = human_pars,
  n_generated = nrow(generated),
  n_quad      = 7
)

cat("PPI++ score lambda:", round(ppi_score$lambda, 3),
    " r =", round(ppi_score$r, 3),
    " N/(n+N) =", round(1 / (1 + ppi_score$r), 3), "\n")
```

A value near zero means the paired LLM responses do not systematically reduce
gradient variance in the person-level score formulation. This is expected for
stochastic binary LLM responses. For F = Y the formula recovers $N/(n+N)$.

## Step 2: Recommended — ability-risk tuning with MML estimator

The key result from the [Linking and Gradient Asymmetry](linking-comparison.html)
vignette: when the LLM parameters differ from human parameters, the frozen
expected-count estimator drives $\lambda \to 0$ due to an artificial gradient
asymmetry. The MML estimator removes this asymmetry. With a good predictor
(F = Y), `tune_lambda_ability_risk` with MML correctly selects $\lambda > 0$.

```{r ability-tuning-mml}
ability_tuned_mml <- tune_lambda_ability_risk(
  lambda_grid  = lambda_grid,
  observed     = observed,
  predicted    = predicted,
  generated    = generated,
  target_resp  = observed,
  initial_pars = human_pars,
  fit_fn       = fit_mixed_subjects_mml,     # marginal MML estimator
  n_quad       = 7,
  control      = list(maxit = 100)
)

ability_tuned_mml$summary[, c("lambda", "mean_param_var", "mean_total_risk",
                               "convergence", "selection_risk")]
ability_tuned_mml$best_lambda
```

`selection_risk` is `Inf` for any candidate with non-zero convergence code or
non-finite risk, protecting selection from numerical failures.

The selected calibration carries a correctly-sized Louis-corrected covariance:

```{r mml-vcov}
Sigma_mml <- vcov(ability_tuned_mml$best_fit)  # dispatches to vcov_mixed_subjects_mml
dim(Sigma_mml)
```

## Step 3: Ability-score risk (inspect components)

```{r risk-components}
risk <- ability_risk(
  resp        = observed,
  fit_or_pars = ability_tuned_mml$best_fit,
  vcov        = Sigma_mml
)
risk$summary
```

`mean_squared_error` is `NA` because `theta_true` was not supplied; it is only
computed in simulation studies. `mean_param_var` is the propagated item-parameter
uncertainty — the criterion that `tune_lambda_ability_risk` minimizes.

In simulation studies, pass `theta_true = theta_human` to also include squared
ability-estimation error:

```{r risk-with-truth}
ability_tuned_truth <- tune_lambda_ability_risk(
  lambda_grid  = lambda_grid,
  observed     = observed,
  predicted    = predicted,
  generated    = generated,
  target_resp  = observed,
  theta_true   = theta_human,
  initial_pars = human_pars,
  fit_fn       = fit_mixed_subjects_mml,
  n_quad       = 7,
  control      = list(maxit = 100)
)

ability_tuned_truth$summary[, c("lambda", "mean_param_var",
                                 "mean_squared_error", "mean_total_risk")]
```

## Step 4: Cross-fitted ability-risk tuning

`tune_lambda_ability_risk_crossfit()` estimates $\lambda$ per fold on training
data, then fits a final split-sample model. Two important arguments:

- `target_mode = "fixed"` (default): the full `target_resp` is used for every
  fold's risk evaluation, which is correct when the target is an operational
  scoring population independent of the calibration sample.
- `target_mode = "row_aligned"`: subsets `target_resp` to training rows, valid
  when `target_resp = observed`.

```{r crossfit}
split_id <- rep(1:2, length.out = nrow(observed))

crossfit_tuned <- tune_lambda_ability_risk_crossfit(
  lambda_grid  = c(0, 0.5, 1),
  observed     = observed,
  predicted    = predicted,
  generated    = generated,
  split_id     = split_id,
  initial_pars = human_pars,
  n_quad       = 7,
  control      = list(maxit = 100)
)

crossfit_tuned$lambda_by_split
```

## Step 5: Frozen expected-count estimator (fast approximation)

The default `fit_fn = fit_mixed_subjects` uses the older frozen expected-count
estimator. This is faster but produces inflated discriminations when the LLM
parameters differ from human parameters, driving $\lambda \to 0$ even when the
LLM is informative.

```{r frozen-ec}
ability_tuned_ec <- tune_lambda_ability_risk(
  lambda_grid  = lambda_grid,
  observed     = observed,
  predicted    = predicted,
  generated    = generated,
  target_resp  = observed,
  initial_pars = human_pars,
  n_quad       = 7,
  slope_upper  = 4,           # required to prevent divergence
  control      = list(maxit = 100)
)

ability_tuned_ec$best_lambda
```

`slope_upper = 4` is required to prevent discriminations from diverging. The
MML estimator does not need this cap because it has no false minimum at large
discrimination values.

## Choosing a procedure

| Procedure | Objective | When to use |
|-----------|-----------|-------------|
| `tune_lambda_ppi_score()` | $\text{Tr}(\Sigma_\gamma)$ | Method diagnostics only |
| `tune_lambda_ability_risk(..., fit_fn = fit_mixed_subjects_mml)` | $E[g'\Sigma_\gamma g]$, Louis bread | **Recommended default** |
| `tune_lambda_ability_risk()` (default `fit_fn`) | Same risk, EM bread | Fast approximation; requires `slope_upper` |
| `tune_lambda_ability_risk_crossfit()` | Out-of-sample risk | Final inferential analyses |
| `tune_lambda_ability_risk_item()` | Per-item risk (approx.) | Experimental; some items poor predictors |

The target population matters. `target_resp = observed` tunes for the observed
human response patterns. In operational scoring, use a larger target matrix
representing the actual scoring population.
