---
title: "Per-Item Lambda (Experimental)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Per-Item Lambda (Experimental)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

> **Experimental.** Per-item lambda uses a frozen expected-count approximation,
> not the full marginal-MML objective. The IRT marginal likelihood integrates the
> *joint* response pattern and does not decompose item-wise, so a theoretically
> principled per-item marginal objective does not yet exist in this package.
> Results from per-item tuning should be treated as approximate and validated
> against the scalar-lambda MML baseline. See the documentation of
> `tune_lambda_ability_risk_item()` for details.

## Why per-item lambda?

A single global $\lambda$ is fragile: if even one item has poorly-correlated
LLM predictions, the ability-risk criterion may force the global optimum to
$\lambda = 0$, preventing all items from benefiting. Per-item $\lambda_j$ allows
each item to draw on the LLM data at its own optimal level.

Consider a test with eight items where:
- Items 1–4 are straightforward factual questions — the LLM predicts these well.
- Items 5–8 require nuanced reasoning — the LLM is nearly random on these.

With a scalar $\lambda$, the four poor items push the optimum toward 0. With
per-item $\lambda_j$, items 1–4 get $\lambda_j \approx 0.5$–0.7 while items 5–8
get $\lambda_j \approx 0$.

## Simulate a heterogeneous test

```{r simulate}
library(mixedsubjectsirt)
library(ggplot2)

set.seed(2026)

n_human    <- 400
n_generated <- 1200
n_items    <- 8
n_good     <- 4   # items where LLM predicts well

true_pars <- data.frame(
  item = paste0("Item", seq_len(n_items)),
  a    = seq(0.8, 1.6, length.out = n_items),
  d    = seq(-1.1, 1.1, length.out = n_items)
)
true_pars$b <- -true_pars$d / true_pars$a

theta_human <- rnorm(n_human)
observed    <- simulate_2pl(theta_human, true_pars)

# LLM: good for items 1–4 (same DGP), poor for 5–8 (random noise)
llm_pars_good <- true_pars
llm_pars_poor <- true_pars
llm_pars_poor$a <- pmax(0.05, rnorm(n_items, 0, 0.1))  # near-random
llm_pars_poor$d <- rnorm(n_items, 0, 2)
llm_pars_poor$b <- -llm_pars_poor$d / llm_pars_poor$a

# Build predicted (same subjects as human)
predicted <- observed   # F = Y for first 4 items
predicted[, 5:8] <- simulate_2pl(theta_human, llm_pars_poor)[, 5:8]

# Build generated
generated_good <- simulate_2pl(rnorm(n_generated), true_pars)
generated_poor <- simulate_2pl(rnorm(n_generated), llm_pars_poor)
generated <- cbind(generated_good[, 1:4], generated_poor[, 5:8])
colnames(generated) <- true_pars$item
```

The first four items have perfect paired predictions (F = Y); items 5–8 have
near-random LLM predictions.

## Step 1: Fit 2PL baseline and get global scalar lambda

```{r baseline}
human_pars <- fit_2pl(observed, technical = list(NCYCLES = 500))$pars

global_tuned <- tune_lambda_ability_risk(
  lambda_grid  = seq(0, 1, by = 0.1),
  observed     = observed,
  predicted    = predicted,
  generated    = generated,
  initial_pars = human_pars,
  fit_fn       = fit_mixed_subjects_mml,
  n_quad       = 11,
  control      = list(maxit = 200)
)

cat("Global scalar best lambda:", global_tuned$best_lambda, "\n")
```

The global scalar is forced to a compromise — the four poor items constrain it
to a value smaller than what items 1–4 could support.

## Step 2: PPI++ score per item (fast diagnostic)

`tune_lambda_ppi_score_item()` applies the Proposition 2 formula independently
per item using the 2×2 diagonal block of $H^{-1}$ and the item-level sub-vectors
of the score matrices. This is fast (no fitting required) and shows which items
are well-predicted.

```{r ppi-item}
ppi_item <- tune_lambda_ppi_score_item(
  observed    = observed,
  predicted   = predicted,
  item_pars   = human_pars,
  n_generated = n_generated,
  n_quad      = 11
)

cat("Per-item PPI++ lambda:\n")
print(data.frame(item = ppi_item$item, lambda = round(ppi_item$lambda, 3)))
cat("N/(n+N) upper bound:", round(n_generated / (n_human + n_generated), 3), "\n")
```

Items 1–4 (F = Y) should show $\lambda_j \approx N/(n+N) = 0.75$; items 5–8
(random LLM) should show $\lambda_j \approx 0$.

## Step 3: Per-item ability-risk tuning

`tune_lambda_ability_risk_item()` uses coordinate descent: for each item $j$, it
finds the $\lambda_j$ in `lambda_grid` that minimises ability-score risk while
holding all other $\lambda_{j'}$ fixed. Starting from the global scalar optimum
(not from all-zeros) is essential — see the note below.

```{r item-risk, cache = FALSE}
item_tuned <- tune_lambda_ability_risk_item(
  lambda_grid  = seq(0, 1, by = 0.25),
  observed     = observed,
  predicted    = predicted,
  generated    = generated,
  initial_pars = human_pars,
  init_lambda  = global_tuned$best_lambda,   # start from global best
  n_quad       = 11,
  n_pass       = 1,
  control      = list(maxit = 200)
)

cat("Per-item ability-risk lambda:\n")
print(data.frame(item = item_tuned$item, lambda = round(item_tuned$lambda, 3)))
```

Items 1–4 should receive positive $\lambda_j$ (good predictor); items 5–8 should
be near zero (poor predictor).

## Step 4: Compare scalar vs. per-item parameter recovery

```{r compare}
fit_scalar   <- global_tuned$best_fit
fit_per_item <- item_tuned$final_fit

rmse <- function(x, y) sqrt(mean((x - y)^2))

comparison <- data.frame(
  item    = true_pars$item,
  true_a  = round(true_pars$a, 3),
  human_a = round(human_pars$a, 3),
  scalar_a = round(fit_scalar$item_pars$a, 3),
  item_a   = if (is.null(fit_per_item)) NA_real_ else
    round(fit_per_item$item_pars$a, 3)
)
knitr::kable(comparison, row.names = FALSE,
  caption = "Discrimination recovery: scalar lambda vs. per-item lambda")

cat("RMSE(a) human-only:   ",
    round(rmse(human_pars$a, true_pars$a), 4), "\n")
cat("RMSE(a) scalar MML:   ",
    round(rmse(fit_scalar$item_pars$a, true_pars$a), 4), "\n")
if (!is.null(fit_per_item)) {
  cat("RMSE(a) per-item MML: ",
      round(rmse(fit_per_item$item_pars$a, true_pars$a), 4), "\n")
}
```

## Important note on initialisation

Starting coordinate descent from all-zeros is not recommended. When all other
items are at $\lambda_j = 0$, each single-item improvement is diluted across the
full ability-risk criterion, making improvements hard to detect. The recommended
workflow is:

1. Fit the scalar global optimum with `tune_lambda_ability_risk(..., fit_fn =
   fit_mixed_subjects_mml)`.
2. Pass that global value as `init_lambda` to `tune_lambda_ability_risk_item()`.
3. The coordinate descent then identifies which items should deviate from the
   global default — raising $\lambda_j$ for well-predicted items and lowering it
   for poorly-predicted ones.

## Approximation caveat

The per-item lambda coordinate descent uses the **frozen Q-function** (not the
full marginal-MML objective) for each candidate evaluation. This is necessary
because the IRT marginal likelihood integrates the joint response pattern and
does not decompose item-wise. The approximation is good when `initial_pars` is
close to the converged parameters. For final reporting, always:

- Compare the per-item fit against the scalar MML baseline.
- Report per-item results as approximate / experimental.
- Use `vcov(item_tuned$final_fit)` for uncertainty (which applies `vcov_mixed_subjects`
  with the vector-lambda bread and meat scaling).
