---
title: "Mixed-Subjects IRT Calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-Subjects IRT Calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

This vignette shows the basic mixed-subjects workflow for a unidimensional 2PL
model. The package expects three response matrices with the same item columns:

- `observed`: human responses.
- `predicted`: LLM responses or probabilities for the same human rows.
- `generated`: additional generated or unlabeled LLM responses.

The fitted objective is

```{r, echo = FALSE, results = "asis"}
cat("`L_human + lambda * (L_generated - L_paired_llm)`")
```

Setting `lambda = 0` gives the human-only expected-count calibration.

## Simulate example data

This example uses simulated LLM responses so the vignette can run without any
external API calls.

```{r simulate}
library(mixedsubjectsirt)
library(ggplot2)

set.seed(2026)

n_human <- 400
n_generated <- 1200
n_items <- 8

true_pars <- data.frame(
  item = paste0("Item", seq_len(n_items)),
  a = seq(0.8, 1.6, length.out = n_items),
  d = seq(-1.1, 1.1, length.out = n_items)
)
true_pars$b <- -true_pars$d / true_pars$a

theta_human <- rnorm(n_human)
observed <- simulate_2pl(theta_human, true_pars)

llm_pars <- true_pars
llm_pars$a <- pmax(0.4, 0.9 * true_pars$a + rnorm(n_items, 0, 0.05))
llm_pars$d <- true_pars$d + 0.25 + rnorm(n_items, 0, 0.15)
llm_pars$b <- -llm_pars$d / llm_pars$a

predicted <- simulate_2pl(theta_human, llm_pars)
generated <- simulate_2pl(rnorm(n_generated, mean = 0.1, sd = 1.05), llm_pars)
```

## Workflow without split samples

The no-split workflow uses all human responses to construct the human posterior
weights and uses those same posterior weights for the paired LLM responses.
That default is controlled by `common_predicted_weights = TRUE`.

```{r no-split}
human_start <- fit_2pl(
  observed,
  technical = list(NCYCLES = 500)
)

human_only <- fit_mixed_subjects(
  observed = observed,
  predicted = predicted,
  generated = generated,
  lambda = 0,
  initial_pars = human_start$pars,
  n_quad = 11
)

mixed_no_split <- fit_mixed_subjects(
  observed = observed,
  predicted = predicted,
  generated = generated,
  lambda = 0.5,
  initial_pars = human_start$pars,
  n_quad = 11
)

mixed_no_split
```

The fitted item parameters are returned in slope-intercept form.

```{r no-split-pars}
head(mixed_no_split$item_pars)
```

## Workflow with split samples

The split workflow cross-fits the posterior weights used for the labeled human
and paired LLM terms. For each split, the initial human 2PL model is estimated
on the other splits, then used to summarize the held-out split. Every human row
still contributes to the final estimating equation once.

```{r split}
mixed_split <- fit_mixed_subjects_split(
  observed = observed,
  predicted = predicted,
  generated = generated,
  lambda = 0.5,
  n_splits = 2,
  seed = 2026,
  n_quad = 11,
  technical = list(NCYCLES = 500)
)

mixed_split
```

## Compare the calibrations

In a real calibration study the true item parameters are unknown. In this
simulation they are useful for checking that the workflow is behaving as
expected.

```{r compare}
estimates <- rbind(
  data.frame(estimator = "human only", human_only$item_pars),
  data.frame(estimator = "mixed no split", mixed_no_split$item_pars),
  data.frame(estimator = "mixed split", mixed_split$item_pars)
)
estimates$true_b <- true_pars$b[match(estimates$item, true_pars$item)]
estimates$true_a <- true_pars$a[match(estimates$item, true_pars$item)]

ggplot(estimates, aes(true_b, b, color = estimator)) +
  geom_abline(slope = 1, intercept = 0, linewidth = 0.4) +
  geom_point(size = 2) +
  labs(x = "True difficulty", y = "Estimated difficulty", color = NULL) +
  theme_minimal()
```

The helper `tune_lambda_grid()` can be used for quick sensitivity checks over
candidate lambda values. The summary reports the optimized mixed objective and
the human expected-count loss, but final lambda selection should be tied to the
study's inferential target or evaluated with a bootstrap/cross-validation plan.

```{r tune}
tuned <- tune_lambda_grid(
  lambda_grid = c(0, 0.25, 0.5, 0.75),
  observed = observed,
  predicted = predicted,
  generated = generated,
  initial_pars = human_start$pars,
  n_quad = 11,
  control = list(maxit = 200)
)

tuned$summary
```
