---
title: "Mixed-Subjects IRT Calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-Subjects IRT Calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

This vignette shows the recommended mixed-subjects workflow for a unidimensional
2PL model using the **marginal maximum-likelihood PPI++ estimator**
(`fit_mixed_subjects_mml`). The package expects three response matrices with the
same item columns:

- `observed`: binary human responses.
- `predicted`: LLM responses or probabilities for the same human rows.
- `generated`: additional generated or unlabeled LLM responses.

The fitted objective is

```{r, echo = FALSE, results = "asis"}
cat(
  "$$L_h^{\\mathrm{marg}}(\\gamma) + \\lambda\\bigl[",
  "L_g^{\\mathrm{marg}}(\\gamma) - L_p^{\\mathrm{marg}}(\\gamma)\\bigr]$$"
)
```

where each $L^\\mathrm{marg}$ is the true IRT marginal negative log-likelihood,
with posteriors recomputed from the current candidate $\\gamma$ at every gradient
step.  Setting $\\lambda = 0$ recovers the human-only MML calibration.  See the
[Choosing Lambda](lambda-tuning.html) vignette for the scientific background on
why the marginal-MML objective is preferred over the older frozen expected-count
estimator.

## Simulate example data

```{r simulate}
library(mixedsubjectsirt)
library(ggplot2)

set.seed(2026)

n_human    <- 400
n_generated <- 1200
n_items    <- 8

true_pars <- data.frame(
  item = paste0("Item", seq_len(n_items)),
  a    = seq(0.8, 1.6, length.out = n_items),
  d    = seq(-1.1, 1.1, length.out = n_items)
)
true_pars$b <- -true_pars$d / true_pars$a

theta_human <- rnorm(n_human)
observed    <- simulate_2pl(theta_human, true_pars)

# LLM with ~10% attenuated discrimination and a small intercept shift
llm_pars <- true_pars
llm_pars$a <- pmax(0.4, 0.9 * true_pars$a + rnorm(n_items, 0, 0.05))
llm_pars$d <- true_pars$d + 0.25 + rnorm(n_items, 0, 0.15)
llm_pars$b <- -llm_pars$d / llm_pars$a

predicted <- simulate_2pl(theta_human, llm_pars)
generated <- simulate_2pl(rnorm(n_generated), llm_pars)
```

## Step 1: Fit the human baseline

```{r human-baseline}
human_start <- fit_2pl(observed, technical = list(NCYCLES = 500))
```

## Step 2: Fit the marginal-MML mixed-subjects model

`fit_mixed_subjects_mml()` recomputes posterior weights at every gradient
evaluation, eliminating the frozen-posterior gradient asymmetry that causes the
older `fit_mixed_subjects()` to inflate item discriminations when the LLM
parameters differ from the human calibration.

```{r mml-fit}
mixed_mml <- fit_mixed_subjects_mml(
  observed  = observed,
  predicted = predicted,
  generated = generated,
  lambda    = 0.5,
  initial_pars = human_start$pars,
  n_quad    = 11
)

mixed_mml
```

```{r mml-pars}
mixed_mml$item_pars
```

## Step 3: Select lambda by ability-score risk

`tune_lambda_ability_risk()` with `fit_fn = fit_mixed_subjects_mml` evaluates
candidates on the MML objective and selects the $\lambda$ that minimises
propagated ability-score risk $E[g'\Sigma_\gamma g]$, where $\Sigma_\gamma$ is
the **Louis-corrected** marginal sandwich covariance.

```{r tune-lambda}
ability_tuned <- tune_lambda_ability_risk(
  lambda_grid  = seq(0, 1, by = 0.2),
  observed     = observed,
  predicted    = predicted,
  generated    = generated,
  target_resp  = observed,
  initial_pars = human_start$pars,
  fit_fn       = fit_mixed_subjects_mml,
  n_quad       = 11,
  control      = list(maxit = 200)
)

ability_tuned$summary
ability_tuned$best_lambda
```

When the LLM parameters differ substantially from the human calibration the
ability-risk criterion correctly selects $\lambda \approx 0$, recovering the
human-only estimate.  The `selection_risk` column shows `Inf` for any candidate
that failed to converge, so the selection is protected against numerical failures.

## Step 4: Inspect the covariance

`vcov()` on a scalar-lambda MML fit automatically uses
`vcov_mixed_subjects_mml()`, which applies Louis' (1982) observed-information
correction — the marginal bread is $H_\mathrm{comp} - I_\mathrm{miss}$ rather
than the EM complete-data Hessian alone.

```{r covariance}
Sigma <- vcov(ability_tuned$best_fit)
dim(Sigma)  # 2J × 2J
```

## Compare calibrations

```{r compare}
human_only <- fit_mixed_subjects_mml(
  observed  = observed,
  predicted = predicted,
  generated = generated,
  lambda    = 0,
  initial_pars = human_start$pars,
  n_quad    = 11
)

estimates <- rbind(
  data.frame(estimator = "human only",  human_only$item_pars),
  data.frame(estimator = "MML lambda = 0.5", mixed_mml$item_pars),
  data.frame(estimator = "MML ability-risk",
             ability_tuned$best_fit$item_pars)
)
estimates$true_b <- true_pars$b[match(estimates$item, true_pars$item)]
estimates$true_a <- true_pars$a[match(estimates$item, true_pars$item)]

ggplot(estimates, aes(true_b, b, colour = estimator)) +
  geom_abline(slope = 1, intercept = 0, linewidth = 0.4) +
  geom_point(size = 2) +
  labs(x = "True difficulty", y = "Estimated difficulty", colour = NULL) +
  theme_minimal()
```

## When the LLM is a good predictor

To verify that the method selects $\lambda > 0$ when the LLM genuinely helps,
simulate with `predicted = observed` (F = Y, the perfect paired predictor):

```{r perfect-predictor}
tuned_fy <- tune_lambda_ability_risk(
  lambda_grid  = seq(0, 1, by = 0.2),
  observed     = observed,
  predicted    = observed,          # F = Y: perfect paired predictor
  generated    = simulate_2pl(rnorm(n_generated), true_pars),
  target_resp  = observed,
  initial_pars = human_start$pars,
  fit_fn       = fit_mixed_subjects_mml,
  n_quad       = 11,
  control      = list(maxit = 200)
)

tuned_fy$best_lambda   # expect > 0, near N/(n+N) = 0.75
tuned_fy$summary[, c("lambda", "mean_param_var", "convergence")]
```

The ability risk decreases from $\lambda = 0$ to a minimum near the theoretical
upper bound $N/(n+N) = 1200/1600 = 0.75$, confirming the estimator correctly
identifies and exploits a highly informative predictor.
