---
title: "Simulation Validation of the Mixed-Subjects MML Estimator"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Simulation Validation of the Mixed-Subjects MML Estimator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)
library(knitr)
```

This vignette presents a simulation study validating the scalar 2PL marginal
maximum-likelihood (MML) PPI++ estimator (`fit_mixed_subjects_mml`) and its
ability-risk tuning (`tune_lambda_ability_risk`). The full reproducible harness
lives in the package's `simulations/` directory; the tables below report results
from runs with the rep counts noted in each section.

The study answers three questions:

1. **Does λ-selection track predictor quality?** (Scenario 1)
2. **Are the Louis-corrected standard errors honest?** (Scenario 2 — the key check)
3. **Does the method improve downstream scoring without ever harming it?** (Scenario 3)

## Design

Human responses come from a true 8-item 2PL model
(`a ∈ [0.8, 1.6]`, `d ∈ [-1.1, 1.1]`) with `n = 400` human subjects and
`N = 1200` generated subjects, abilities drawn from a standard normal. Five
predictor regimes vary the quality of the paired LLM predictions `F`:

```{r design-table}
design <- data.frame(
  Regime = c("R1 perfect (F=Y)", "R2 conditional mean", "R3 same-DGP draw",
             "R4 independent noise", "R5 LLM shift"),
  `Predicted F` = c(
    "predicted = observed (exact)",
    "true probabilities p(theta) (fractional)",
    "fresh draw from the true model",
    "draw from scrambled item parameters",
    "draw from attenuated/shifted parameters"),
  `Predictor quality` = c("perfect", "oracle (but see below)", "modest",
                          "none", "low"),
  check.names = FALSE
)
kable(design, caption = "Five predictor regimes, ordered best to worst.")
```

The control parameter λ ∈ [0, 1] sets how strongly the generated LLM sample is
used after subtracting the paired LLM correction;
λ = 0 is the human-only calibration.

## Scenario 1: λ selection (100 reps)

For each regime we tune λ by ability-score risk and record the selected value.
We also report the theoretical PPI++-score λ for reference.

```{r lambda-table}
lam <- data.frame(
  Regime      = c("R1 perfect (F=Y)", "R2 conditional mean", "R3 same-DGP draw",
                  "R4 independent noise", "R5 LLM shift"),
  `mean lambda` = c(0.750, 0.028, 0.108, 0.032, 0.102),
  median        = c(0.75, 0.00, 0.10, 0.00, 0.10),
  `prop lambda=0` = c(0.00, 0.96, 0.26, 0.74, 0.31),
  `PPI score lambda` = c(0.750, 0.016, 0.004, 0.000, 0.002),
  check.names = FALSE
)
kable(lam, caption = "Selected lambda by regime (ability-risk tuning, 100 reps).")
```

**Findings.**

- **R1 perfect → λ = 0.750**, exactly `N / (n + N) = 1200 / 1600`, the theoretical
  maximum for a perfect paired predictor. The PPI++-score λ agrees to the decimal.
- **R4 independent noise → λ ≈ 0** (74% of reps select 0): a useless predictor is
  correctly rejected.
- **R3 same-DGP and R5 LLM shift → λ ≈ 0.10**: fresh real responses (R3) and a
  correlated-but-biased LLM (R5) each carry a little signal, and the criterion uses
  a little of it.
- **R2 conditional mean → λ ≈ 0** (96% of reps select 0). This is the one
  counterintuitive result and is discussed next.

### Why the "oracle" predictor (R2) is declined

R2 feeds the method the exact true probability of a correct response for each
person — seemingly the best possible predictor. Yet λ is almost always 0. The
reason: PPI++ extracts information from the **residuals** between actual answers
and the model's expectation. A real 0/1 response carries a residual; the
conditional-mean prediction *is* the expectation, so its residual is zero and it
carries nothing to correct with. A *noisier* predictor that emits actual responses
(R3) is therefore more useful than a *perfect* one that emits smooth probabilities
(R2). The practical lesson for applications: to benefit from synthetic/LLM data,
have it produce sampled responses, not averaged probabilities.

## Scenario 2: Louis-corrected SE coverage (200 reps) — the key check

We fix λ = 0.5 in the two regimes where the combined estimator is consistent for
the true parameters (R1 and R3), and compare empirical coverage of Wald intervals
built from two covariance estimates of the *same* fit: the Louis-corrected
marginal sandwich (`vcov()` dispatch) and the EM complete-data Hessian bread
(`vcov_mixed_subjects()`, a deliberate bypass).

```{r coverage-table}
cov <- data.frame(
  Regime = c("R1 perfect (F=Y)", "R3 same-DGP draw"),
  `Louis 90%` = c(0.909, 0.916),
  `Louis 95%` = c(0.955, 0.957),
  `EM 90%`    = c(0.713, 0.727),
  `EM 95%`    = c(0.787, 0.797),
  `SE ratio (Louis/EM)` = c(1.63, 1.62),
  check.names = FALSE
)
kable(cov, caption = paste("Item-parameter CI coverage (200 reps).",
  "Nominal targets 0.90 and 0.95."))
```

**The Louis correction restores nominal coverage; the EM bread under-covers.**
Louis intervals attain 0.909/0.955 and 0.916/0.957 against nominal 0.90/0.95,
while the EM bread covers only ~0.71/0.79 — it understates uncertainty because it
ignores the missing information about each subject's latent ability. The Louis
standard errors are ~1.6× wider, which is the difference between an honest interval
and an overconfident one.

Coverage broken down by parameter type confirms the correction works uniformly:

```{r coverage-partype}
cp <- data.frame(
  `Parameter` = c("discrimination a", "intercept d"),
  `Louis 90%` = c(0.91, 0.91),
  `Louis 95%` = c(0.96, 0.95),
  `EM 90%`    = c(0.60, 0.84),
  check.names = FALSE
)
kable(cp, caption = "Coverage by parameter type (averaged over R1 and R3).")
```

### A bug this scenario caught

An earlier version of this scenario showed near-zero coverage for discrimination
while intercept coverage was nominal. That pattern — a uniform downward bias in all
slopes with unbiased intercepts — is the signature of a latent-scale error, and it
traced to `make_quadrature()` building an N(0, 2) ability grid instead of N(0, 1)
(a spurious √2 rescaling of standard-normal Gauss–Hermite nodes). Every
discrimination in the package had been biased downward by ~1/√2 ≈ 0.71. The bug was
invisible to relative checks (λ-selection), to gradient finite-difference checks
(loss and gradient were mutually consistent), and to the standard-error calibration
itself (the SEs correctly described the *biased* estimator). Only coverage against
ground truth exposed it. After the fix, discrimination coverage rose from 0.04 to
0.91. This is the strongest argument for ground-truth simulation validation: it
catches errors that internal consistency checks cannot.

## Scenario 3: downstream payoff and no-harm (100 reps)

We score a held-out sample of 1000 subjects with known ability and compare
ability-score RMSE for the human-only calibration (λ = 0) and the tuned MML
calibration.

```{r downstream-table}
dwn <- data.frame(
  Regime = c("R1 perfect (F=Y)", "R2 conditional mean", "R3 same-DGP draw",
             "R4 independent noise", "R5 LLM shift"),
  `mean lambda` = c(0.750, 0.000, 0.108, 0.032, 0.102),
  `RMSE human`  = c(1.4961, 1.4958, 1.5020, 1.5064, 1.4994),
  `RMSE tuned`  = c(1.4921, 1.4958, 1.5010, 1.5060, 1.4981),
  `delta (tuned-human)` = c(-0.0040, 0.0000, -0.0010, -0.0004, -0.0013),
  `bias a`      = c(0.011, 0.019, 0.038, 0.005, 0.031),
  check.names = FALSE
)
kable(dwn, caption = "Downstream ability-score RMSE (100 reps).")
```

**Findings.**

- **No harm anywhere.** The tuned RMSE is never worse than human-only
  (delta ≤ 0 in every regime). When the predictor is useless (R2, R4) the method
  falls back to the human-only estimate; when it helps (R1, R3, R5) it improves,
  modestly.
- **Discrimination is unbiased** at the selected λ (|bias_a| ≤ 0.04 everywhere),
  confirming the quadrature fix and that the tuner is not selecting degenerate fits.
- **The RMSE gains are small** because an 8-item test is measurement-limited:
  ability scoring is dominated by the irreducible measurement error of a short
  test, so even a perfect predictor (R1) moves the scoring RMSE only from 1.4961 to
  1.4921. The method's larger payoff is in **item-parameter precision** — visible
  in the Scenario 2 standard errors — which matters most when calibration
  uncertainty propagates into operational use.

### A robustness guard

Building this scenario surfaced a failure mode: with the fractional R2 predictor at
large λ, the MML objective has a "converged" minimum at runaway discrimination
(a ≈ 1000). Such a fit reports huge item information, which collapses its covariance
and makes the model-based ability risk spuriously small — so the tuner could select
it even though it scores worse on held-out data. The ability-risk tuners now reject
any candidate whose maximum discrimination exceeds `max_discrimination` (default
10); a 2PL discrimination above that is never a real calibration. With the guard,
all R2 reps correctly select λ = 0.

## Summary

| Question | Result |
|----------|--------|
| Does λ track predictor quality? | Yes: perfect → N/(n+N), noise → 0, intermediate → small positive |
| Are the Louis SEs honest? | Yes: nominal coverage for both `a` and `d`; the EM bread under-covers |
| Does the method ever harm scoring? | No: tuned RMSE ≤ human-only in every regime |

The headline practical guidance: use `fit_mixed_subjects_mml` with
`tune_lambda_ability_risk`, rely on `vcov()` (Louis-corrected) for inference, have
synthetic data emit sampled responses rather than probabilities, and treat the
conditional-mean and runaway-discrimination behaviours as the two documented edge
cases.

## Reproducing these results

From the package root:

```r
# rep counts used above; second argument is the number of cores
# Rscript simulations/run_lambda_selection.R 100 8
# Rscript simulations/run_coverage.R 200 8
# Rscript simulations/run_downstream.R 100 8
# Rscript simulations/figures.R
```

See `simulations/README.md` for the full design, the deterministic per-task
seeding (results are identical serially or in parallel), and interpretation notes.
